\name: A flexible and data-driven K-process model for nucleosynthesis

### 2024-10-25 2024_KPM_paper/LaTex/main.tex

\documentclass[modern, linenumbers]{aastex631}
\usepackage{amsmath}
\usepackage{microtype}
\usepackage{soul}
% \usepackage{lineno}
% \linenumbers

@@

\newcommand{\ejg}[1]{\textcolor{red}{EJG says: #1}}
\newcommand{\add}[1]{\textcolor{teal}{#1}}

\newcommand{\hogg}[1]{\textcolor{blue}{Hogg says: #1}}

@@

\author[0000-0003-1124-7378]{Bridget Ratcilffe}
\affiliation{Leibniz-Institut f\"{u}r Astrophysik Potsdam (AIP), An der Sternwarte 16, 
14482 Potsdam, Germany}

@@

\begin{abstract}\noindent
The element abundance pattern found by APOGEE in Milky Way disk stars is close to two-dimensional, dominated by production from one prompt process and one delayed process.
This is remarkable, since the elements are known to be produced by many more than two nucleosynthetic sources.
We fit the abundances of 14 elements for 48,659 red-giant stars from APOGEE DR17 using a flexible, data-driven $K$-process model---dubbed \name.
In our fiducial model, with $K=2$, each abundance in each star is described as the sum of a prompt and a delayed process contribution.
We find that \name{} with $K=2$ is able to explain the abundances well, recover the observed abundance bimodality, and detect the bimodality over a greater range in metallicity than previously has been possible.
We compare to prior work by \citet{weinberg2022}, finding that \name{} produces similar results, but that \name{} better predicts stellar abundances, especially for elements C+N and Mn and for stars at super-solar metallicities.
The model makes assumptions, including especially that it fixes some parameters to break degeneracies and improve interpretability; we find that some of the nucleosynthetic implications are dependent upon these detailed parameter choices.
We add a third and fourth process (to make $K=4$), finding that the additional processes give the model more freedom and improve the model's ability to predict the stellar abundances, as expected, but they don't qualitatively change the story. 
The results of \name{} have implications for the formation of the Galaxy disk, the relationship between abundances and ages, and the physics of nucleosynthesis.
\end{abstract}

@@

At the present day, stellar surface abundances are not very well explained by purely \textsl{ab initio}, physics-driven models.
Theoretical yields vary from data set to data set, as they are dependent on progenitor properties and explosion assumptions \citep[e.g.,][]{rybizki2017, blancato2019, buck2021, griffith2021b}. 
The wide parameter space of progenitor and supernova models coupled with uncertainties in reaction rates and explosion physics hinder the creation of an accurate nucleosynthetic model from theory alone.
In the long run, it is incumbent upon us to understand these issues and correct the assumptions or calculations underlying our nucleosynthetic and astrophysical models.
In the short run, however, we gather data---tens of millions of abundance measurements on millions of stars in different astronomical surveys such as RAVE, SEGUE, LAMOST, Gaia-ESO, APOGEE/MWM, GALAH, and H3 \citep{steinmetz2006, yanny2009, gilmore2012, desilva2015, luo2015, majewski2017, conroy2019}.
This raises the question: \emph{Can we take a data-driven approach to nucleosynthesis?}

@@

We build upon the work of \citet[][hereafter G22]{griffith2019, griffith2022} and \citet[][hereafter W19, W22]{weinberg2019, weinberg2022}, who used the bimodality in [Mg/Fe] vs. [Fe/H]\footnote{Where $[{\rm X}/{\rm Y}] = \log({\rm X}/{\rm Y}) - \log({\rm X}/{\rm Y})_{\odot}$ and $({\rm X}/{\rm Y})_{\odot}$ is the solar abundance ratio.} \citep[e.g.,][]{fuhrmann1998, bensby2003, adibekyan2012} to separate stars into populations with high and low SNIa enrichment. Using the median [X/Mg] vs. [Mg/H] abundance trends, these works explain data from the GALAH\footnote{GALAH = GALactic Archaeology with HERMES.} and SDSS-IV APOGEE\footnote{APOGEE = Apache Point Observatory Galactic Evolution Experiment, part of the Sloan Digital Sky Survey} surveys, respectively, with a two-process model. Because the median abundance trends in [X/Mg] vs. [Mg/H] space are largely insensitive to aspects of chemical evolution, such as outflows and variations in star formation history (W19), the population abundance trends are set by the nucleosynthetic processes and can be used to empirically constrain Galactic enrichment.

@@

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data.
In the $K$ Process Model (\name{}), we find the intersection between reliable facts about nucleosynthesis and good abundance measurements to build an edifice of Galactic enrichment.
The model is hierarchical, in that it learns some parameters (process vectors) that are shared across all stars, but different for each element, and some parameters (process amplitudes) that are shared across all elements, but different for each star.
The parameters output by our model can thus be used as de-noised abundance labels; these will sharpen relationships between abundances and stellar parameters (including birth location and time). Our main contribution is to construct a data-driven model for nucleosynthesis that has good statistical properties, only enforcing constraints to break rotational degeneracies. All other \name{} parameters are set by the data with no fixed normalization.

@@

As in W22 and G22, we propose that all stellar abundances can be generated by a combination of $K$ nucleosynthetic processes.
In this picture, each element has $K$ metallicity-dependent process vector components that are shared across the full stellar sample, while each star individually has $K$ process amplitudes, which apply across all elements, such that the expected logarithmic abundance of element $j$ relative to H in star $i$ ($m_{ij}$) is defined as:
\begin{equation}\label{eq:mij_k}
    m_{ij} = \log_{10} \, \sum^K_{k=1} A_i^k \, q_{k,j}^Z ~.
\end{equation}
Each star $i$ has $K$ process amplitudes ($A^k_i$)
and each element $j$ has $K$ metallicity dependent process vector components ($q_{k,j}^{Z}$). 
The $Z$ superscript denotes the dependence of the process vectors on metallicity, $Z$, taken to be [Mg/H].
The observed abundance can be expressed
\begin{equation}\label{eq:xh}
    [\text{X}_j/\text{H}]_i = m_{ij} + \text{noise},
\end{equation}
where ``noise'' represents observational noise and/or other sources of intrinsic abundance scatter that are not included in this model. For detailed examples of a similar model with $K=2$, see Section 2 and Figures 2 and 3 of W22.

In \name{}, we put forth the following set of assumptions:

@@

\paragraph{2. Linearity}
At every metallicity, the (linear) (X/H) abundances of a star can be expressed as a linear combination of $K$ processes.
These processes themselves will depend on metallicity, but a linear sum is sufficient to explain all element abundances at any overall metallicity.
Because yields depend on stellar masses and progenitor structure, and because different stars can get to their metallicities by different histories, this assumption must be at least slightly wrong in detail.

@@

\paragraph{4. Mg production}
All Mg is produced in a prompt process and no other processes contribute to its production.
This is substantiated by theoretical yields where Mg is purely produced by prompt CCSN \citep[e.g.,][]{anderson2019, rybizki2017}.
This assumption (along with non-negativity) breaks a set of symmetries in the process space and makes the processes quasi-interpretable in terms of nucleosynthesis sources.
Because such a prompt process is likely dominated by CCSN \citep[e.g.,][]{andrews2017}, we label the first process with ``CC''.

@@

\paragraph{5. Fe production}
Fe is produced through a combination of a prompt and delayed process. Because the delayed process is likely dominated by SNIa \citep[e.g.,][]{andrews2017}, we label the delayed process with ``Ia''\footnote{While other enrichment channels with similar timescales may be included in the respective processes, the ``CC'' and ``Ia'' naming convention conforms to the choices in W22 and G22, and avoids the possible confusion of process numbers (1 and 2) with supernova type (II and Ia).}.
While the prompt process constraint (Mg) is grounded in nucleosynthesis theory, there is no equivalent nucleosynthesis fact to constrain the delayed process.
To break model degeneracies, we also fix the Fe process vector components such that

@@

\paragraph{6. Metallicity dependence}
We permit the process vector components for all elements other than Mg and Fe to float as a function of metallicity.
The variation is parameterized by a linear spline in log-process space, attached to a set of variable control points. We assume that a particular set of hard-coded knots are sufficient to capture the metallicity dependence.

\paragraph{7. APOGEE abundances and uncertainties}
We assume that the APOGEE abundances and uncertainties can be used for this project.
This is not the same as assuming that they are correct, but rather that it is possible and useful to build an interpretable model to explain them. We describe the potential data systematics in Section~\ref{sec:data}.
For our purposes, we care mainly about the statistical observational errors rather than systematics that arise from imperfect modeling of the spectra such as NLTE effects, though differential systematics across the sample can artificially add abundance scatter. The actual derived values of $q_{k,j}^{Z}$ will be affected by systematic offsets in the abundances. We add a softening parameter $Q$ (Equation~\ref{eq:inflate_ivar}) to allow for the possibility that APOGEE observational errors are underestimated or that there is intrinsic scatter around the \name{} predictions.

@@

Because we don't want to be too drawn or influenced by outlier points, we don't use the observed errors $\sigma_{{\rm obs},ij}$ in the likelihood, but instead we soften them in the spirit of iteratively reweighted least squares (\citealt{irls}):
\begin{equation}\label{eq:inflate_ivar}
    \frac{1}{\sigma_{ij}^2} = \frac{Q^2/\sigma_{{\rm obs},ij}^2}{Q^2 + (\xh_{ij} - m_{ij})^2 /\sigma_{{\rm obs},ij}^2 },    
\end{equation}
where $Q$ is a softening parameter. Our results are largely insensitive to the choice of $Q$ (e.g., $1 < Q < 10$), so we choose to set $Q=5$. Very small values (e.g., $Q=.1$) will erase some of the abundance structure and produce poorer fits.

@@

These cuts result in a sample of 48,659 stars that span the Galactic disk. We plot their $Z$ vs. $R$ location, as well as the distributions of distances and eccentricities in Figure~\ref{fig:star_dist}, taking distances and kinematics from \citep{queiroz2023}. While our stellar sample extends from Galactic center, to the outer disk, to the halo, the majority of our stars ($75\%$) are within 3.5 kpc of the sun. Further, $94\%$ of our stellar sample has an eccentricity less than 0.4, indicative of \textit{in situ} origin. In this paper, we assume that the KPM fits will be consistent across the Galactic disk, as the median high-Ia and low-Ia [X/Mg] vs. [Mg/H] abundance tends are insensitive to Galactic location (W19, \citealp{griffith2021a}).  
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/star_sample.pdf}
    \caption{Left: distribution of our stellar sample in $Z$ (kpc) vs. $R$ (kpc) where (0,0) is the Galactic center. Center: distribution of stellar distances (kpc). Right: distribution of stellar eccentricity. Our stellar sample spans the Galactic disk, but the majority of our stars are within 3.5 kpc of the sun and have kinematics consistent with \textit{in situ} origin.}
    \label{fig:star_dist}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/all_param1.pdf}
    \caption{Abundance distributions and \name{} parameters for C+N, $\alpha$, and light odd-$Z$ elements. First column: observed abundance distributions in [X/Mg] vs. [Mg/H]. Second column: The predicted [X/Mg] vs. [Mg/H] abundance distribution of the fiducial model plus estimated noise. By comparing the first two columns we can evaluate the success of \name{} in reproducing the observed abundance distributions. Third column: process vector components $\qcc$ (thin, purple) and $\qIa$ (thick, orange) from this work (G23, solid, dark lines) and W22 (light, dashed lines). Overall offsets between the solid and dashed lines are driven largely by our normalization that places the [Mg/Fe] plateau at $+0.4$, rather than $+0.3$ in W22. Fourth column: distribution of fractional contribution from the prompt process ($\fcc$) predicted by the fiducial model. We plot the median $\fcc$ values of the low-Ia (orange square) and high-Ia (purple circle) populations in the solar metallicity bin from W22 for comparison. All density plots are logarithmically scaled.}
    \label{fig:all_param1}
\end{figure*}

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/all_param2.pdf}
    \caption{Same as Figure~\ref{fig:all_param1}, but for Fe-peak elements and Ce.}
    \label{fig:all_param2}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/As.pdf}
    \caption{Left: distribution of $\AIa/\Acc$ vs. $\Acc$ for the fiducial model. This plot is similar to a [Fe/Mg] vs. [Mg/H] distribution where $\AIa/\Acc$ is a proxy for [Fe/Mg] and $\Acc$ is a proxy for [Mg/H]. Note that $\Acc = 1$ corresponds to $\mgh = 0$. Center: distribution of $\AIa/\Acc$ vs. $\mgh$. In the left and center panels, we can clearly see the bimodality to high values of $\Acc$ and $\mgh$. Both density plots are logarithmically scaled. Right: distribution of $\AIa/\Acc$ for ranges of [Mg/H] with metallicity bin increasing from top to bottom. [Mg/H] bins are of width 0.325 dex and span $-0.75$ to 0.55.}
    \label{fig:As}
\end{figure*}

@@

Overall, the fiducial model successfully reproduces the observed abundance distributions. It is capable of capturing metalicity dependences and bimodality. The predicted abundances plus estimated noise are not, however, able to reproduce the observed abundance scatter. This is especially noticeable for O, C+N, Na, Al, K, Co, and Ce. For these elements the scatter in the observed abundance distribution is larger than in the predicted distribution, suggesting that the APOGEE observational scatter is underestimated, that there are $\teff$ or $\logg$ dependent abundance trends (e.g., \citealp{griffith2021a}, W22), or that the $K=2$ model is insufficient--a likely case for elements produced by AGB stars, such as C+N and Ce.

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/comp_w22.pdf}
    \caption{Left: cumulative distribution of $\log_{10}(\chi^2)$ for W22 (dashed orange line) and our fiducial model (G23, solid purple line). Right: $\chi^2$ per element for the same model fits with elements ordered by atomic number. Overall, our model has a smaller cumulative $\chi^2$ than the previous two-process model and better predicts the abundances of C+N, Na, K, Mn, Co, Ni, and Ce.}
    \label{fig:comp_w22}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/qccFe_FeMg.pdf}
    \caption{Minimum [Fe/Mg] value attainable for $\qccFe$ and $\dqccFe$ assumptions. Left: for $\dqccFe=0$ with $\qccFe=0.5$, 0.45, 0.4, and 0.35 (light to dark purple). Right: for $\dqccFe=0.15$ with $\qccFe=0.5$, 0.4, 0.35 (light to dark orange).}
    \label{fig:qccFe_FeMg}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/qccFe_FeMgpred.pdf}
    \caption{Top: observed [Fe/Mg] vs. [Mg/H] abundance distribution. Bottom: predicted [Fe/Mg] vs. [Mg/H] abundance distributions plus estimated noise for models with $\dqccFe=0.0$ (middle row) and $\dqccFe=0.15$ (bottom row). The $\qccFe$ value increase from left to right from 0.35 to 0.5. All density plots are logarithmically scaled. Note that the center panels have different $\qccFe$ values (middle: 0.45, bottom: 0.4) and that the predicted abundance distribution for $\qccFe=0.4$ and $\dqccFe=0.0$ can be found in Figure~\ref{fig:all_param2}. The $\qccFe$ and $\dqccFe$ parameters are listed in each panel. By comparing the lower panels to the observed distribution (top) we can evaluate the success or failure of each model; the low-Ia population is most sensitive to the Fe assumptions.}
    \label{fig:qccFe_FeMgpred}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/qccFe_fcc.pdf}
    \caption{Elemental median values of $\fcc$ at solar metallicity for the low-Ia population for \name{} with $\qccFe=0.35$ (darkest purple) to 0.5 (lightest purple) and $\dqccFe=0.0$. Elements are ordered by atomic number. The median $\fcc$ changes most dramatically for elements with strong delayed contributions.}
    \label{fig:qccFe_fcc}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/A4s.pdf}
    \caption{$A^k_i/\Acc$ vs. $\Acc$ distribution for $\AIa$ (left), $A^3_i$ (center), and $A^4_i$ (right). All density plots are logarithmically scaled. The third and fourth processes contribute most to low metallicity stars.}
    \label{fig:A4}
\end{figure*}

>>>>

\documentclass[modern]{aastex631}
\usepackage{amsmath}
\usepackage{microtype}
\usepackage{soul}

@@

\newcommand{\ejg}[1]{\textcolor{red}{EJG says: #1}}
\newcommand{\add}[1]{\textcolor{teal}{#1}}
\newcommand{\remove}[1]{\textcolor{red}{#1}}
\newcommand{\hogg}[1]{\textcolor{blue}{Hogg says: #1}}

@@

\author[0000-0003-1124-7378]{Bridget Ratcliffe}
\affiliation{Leibniz-Institut f\"{u}r Astrophysik Potsdam (AIP), An der Sternwarte 16, 
14482 Potsdam, Germany}

@@

\begin{abstract}\noindent
The element abundance pattern found in Milky Way disk stars is close to two-dimensional, dominated by production from one prompt process and one delayed process. 
This simplicity is remarkable, since the elements are produced by a multitude of nucleosynthesis mechanisms operating in stars with a wide range of progenitor masses. 
We fit the abundances of 14 elements for 48,659 red-giant stars from APOGEE DR17 using a flexible, data-driven K-process model---dubbed \name. 
In our fiducial model, with $K=2$, each abundance in each star is described as the sum of a prompt and a delayed process contribution. 
We find that \name with $K=2$ is able to explain the abundances well, recover the observed abundance bimodality, and detect the bimodality over a greater range in metallicity than previously has been possible. 
We compare to prior work by \citet{weinberg2022}, finding that \name produces similar results, but that \name better predicts stellar abundances, especially for elements C+N and Mn and for stars at super-solar metallicities. 
The model fixes the relative contribution of the prompt and delayed process to two elements to break degeneracies and improve interpretability; we find that some of the nucleosynthetic implications are dependent upon these detailed choices. 
We find that moving to four processes adds flexibility and improves the model's ability to predict the stellar abundances, but doesn't qualitatively change the story. 
The results of \name will help us to interpret and constrain the formation of the Galaxy disk, the relationship between abundances and ages, and the physics of nucleosynthesis.
\end{abstract}

@@

At the present day, stellar surface abundances are not very well explained by purely \textsl{ab initio}, physics-driven models.
Theoretical yields vary from data set to data set, as they are dependent on progenitor properties and explosion assumptions \citep[e.g.,][]{rybizki2017, blancato2019, buck2021, griffith2021b}. 
The wide parameter space of progenitor and supernova models coupled with uncertainties in reaction rates and explosion physics hinder the creation of an accurate nucleosynthetic model from theory alone.
In the long run, it is incumbent upon us to understand these issues and correct the assumptions or calculations underlying our nucleosynthetic and astrophysical models.
In the short run, however, we gather data---tens of millions of abundance measurements on millions of stars in different astronomical surveys such as RAVE \citep{steinmetz2006}, SEGUE \citep{yanny2009}, LAMOST \citep{luo2015}, Gaia-ESO \citep{gilmore2012, gilmore2022}, APOGEE/MWM \citep{majewski2017}, GALAH \citep{desilva2015}, and H3 \citep{conroy2019}.
This raises the question: \emph{Can we take a data-driven approach to nucleosynthesis?}

@@

We build upon the work of \citet[][hereafter G22]{griffith2019, griffith2022} and \citet[][hereafter W19, W22]{weinberg2019, weinberg2022}, who used the bimodality in [Mg/Fe] vs. [Fe/H]\footnote{Where $[{\rm X}/{\rm Y}] = \log({\rm X}/{\rm Y}) - \log({\rm X}/{\rm Y})_{\odot}$ and $({\rm X}/{\rm Y})_{\odot}$ is the solar abundance ratio.} \citep[e.g.,][]{fuhrmann1998, bensby2003, adibekyan2012} to separate stars into populations with high and low SNIa enrichment. As established in \citet{griffith2019}, these populations are referred to as high-Ia and low-Ia, to reflect their enrichment origins, instead of the traditional low-$\alpha$ and high-$\alpha$ nomenclature. We adopt this updated naming convention in this \documentname. Using the median [X/Mg] vs. [Mg/H] abundance trends, these prior works explain data from the GALAH\footnote{GALAH = GALactic Archaeology with HERMES.} and SDSS-IV APOGEE\footnote{APOGEE = Apache Point Observatory Galactic Evolution Experiment, part of the Sloan Digital Sky Survey} surveys, respectively, with a two-process model. Because the median abundance trends in [X/Mg] vs. [Mg/H] space are largely insensitive to aspects of chemical evolution, such as outflows and variations in star formation history (W19), the population abundance trends are set by the nucleosynthetic processes and can be used to empirically constrain Galactic enrichment.

@@

In this work, our main innovations are to relax the assumptions made in G22 and W22, to be more agnostic about the nucleosynthetic processes, and to be more principled with the measurements or inferences from data.
In the $K$ Process Model (\name{}), we find the intersection between reliable facts about nucleosynthesis and good abundance measurements to build an edifice of Galactic enrichment.
The model is hierarchical, in that it learns some parameters (process vectors) that are shared across all stars, but different for each element, and some parameters (process amplitudes) that are shared across all elements, but different for each star.
The parameters output by our model can thus be used as de-noised abundance labels; these will sharpen relationships between abundances and stellar parameters (including birth location and time). Our main contribution is to construct a data-driven model for nucleosynthesis that has good statistical properties, only enforcing a small number of constraints to break the degeneracies that arise in models of this form. All other \name{} parameters are set by the data with no fixed normalization.

@@

As in W22 and G22, we propose that all stellar abundances can be generated by a combination of $K$ nucleosynthetic processes.
In this picture, each element has $K$ metallicity-dependent process vector components that are shared across the full stellar sample, while each star individually has $K$ process amplitudes, which apply across all elements, such that the expected logarithmic abundance of element $j$ relative to H in star $i$ ($m_{ij}$) is defined as:
\begin{equation}\label{eq:mij_k}
    m_{ij} = \log_{10} \, \sum^K_{k=1} A_i^k \, q_{k,j}^Z ~.
\end{equation}
Each star $i$ has $K$ process amplitudes ($A^k_i$)
and each element $j$ has $K$ metallicity dependent process vector components ($q_{k,j}^{Z}$). 
The $Z$ superscript denotes the dependence of the process vectors on metallicity, $Z$, taken to be [Mg/H].
The observed abundance can be expressed
\begin{equation}\label{eq:xh}
    [\text{X}_j/\text{H}]_i = m_{ij} + \text{noise},
\end{equation}
where ``noise'' represents observational noise and/or other sources of intrinsic abundance scatter that are not included in this model. For detailed examples of a similar model with $K=2$, see Section 2 and Figures 2 and 3 of W22, where they demonstrate the vector addition and describe the process parameters for a few example stars.

In \name{}, we adopt the following set of assumptions:

@@

\paragraph{2. Linearity}
At every metallicity, the (linear) (X/H) abundances of a star can be expressed as a linear combination of $K$ processes.
These processes themselves will depend on metallicity, but a linear sum is sufficient to explain all element abundances at any overall metallicity.
Because different stars can get to their metallicities by different histories, and because detailed abundances beyond metallicity must matter at some level, the true enrichment mechanism is at least slightly nonlinear; thus this assumption
must be at least slightly wrong in detail.

@@

\paragraph{4. Mg production}
All Mg is produced in a prompt process and no other processes contribute to its production.
This is substantiated by theoretical yields where Mg is purely produced by prompt CCSN \citep[e.g.,][]{woosley1995, arnett1996, anderson2019, rybizki2017}.
This assumption (along with non-negativity) breaks a set of symmetries in the process space and makes the processes quasi-interpretable in terms of nucleosynthesis sources.
Because such a prompt process is likely dominated by CCSN \citep[e.g.,][]{andrews2017}, we label the first process with ``CC''.

@@

\paragraph{5. Fe production}
Fe is produced through a combination of a prompt and delayed process. Because the delayed process is likely dominated by SNIa \citep[e.g.,][]{thielemann2002, andrews2017}, we label the delayed process with ``Ia''\footnote{While other enrichment channels with similar timescales may be included in the respective processes, the ``CC'' and ``Ia'' naming convention conforms to the choices in W22 and G22, and avoids the possible confusion of process numbers (1 and 2) with supernova type (II and Ia).}.
While the prompt process constraint (Mg) is grounded in nucleosynthesis theory, there is no equivalent nucleosynthesis fact to constrain the delayed process.
To break model degeneracies, we also fix the Fe process vector components such that

@@

\paragraph{6. Metallicity dependence}
We permit the process vector components for all elements other than Mg and Fe to float as a function of metallicity.
The variation is parameterized by a linear spline in log-process space, attached to a set of variable control points, knots, where the piecewise functions are joined. We assume that a particular set of 11 hard-coded knots between [Mg/H] of -0.8 and 0.6 are sufficient to capture the metallicity dependence. We choose knot number and location such that we capture the complex metallicity dependence of the abundance trends while maintaining a sufficient number of stars to fit with each linear component.

\paragraph{7. APOGEE abundances and uncertainties}
We assume that the APOGEE abundances and uncertainties can be used for this project.
This is not the same as assuming that they are correct, but rather that it is possible and useful to build an interpretable model to explain them. We describe the potential data systematics in Section~\ref{sec:data}.
For our purposes, we care mainly about the statistical observational errors rather than systematics that arise from imperfect modeling of the spectra such as NLTE effects, though differential systematics across the sample can artificially add abundance scatter. The actual derived values of $q_{k,j}^{Z}$ will be affected by systematic offsets in the abundances. We add a softening parameter $Q$ (Equation~\ref{eq:inflate_ivar}) to allow for the possibility that APOGEE observational errors are underestimated, or that there is intrinsic scatter around the \name{} predictions.

@@

Because we don't want to be too drawn or influenced by outlier points, we don't use the observed errors $\sigma_{{\rm obs},ij}$ in the likelihood, but instead we soften them in the spirit of iteratively reweighted least squares (\citealt{irls}):
\begin{equation}\label{eq:inflate_ivar}
    \frac{1}{\sigma_{ij}^2} = \frac{Q^2/\sigma_{{\rm obs},ij}^2}{Q^2 + (\xh_{ij} - m_{ij})^2 /\sigma_{{\rm obs},ij}^2 },    
\end{equation}
where $Q$ is a softening parameter. Our results are largely insensitive to the choice of $Q$. We find that the predicted abundances of all elements change by less than 0.01 dex for Q between 1 and 10, so we choose to set $Q=5$. Very small Q values (e.g., $Q=.1$) will erase some of the abundance structure and produce poorer fits.

@@

These cuts result in a sample of 48,659 stars that span the Galactic disk. We plot their $Z$ vs. $R$ location, as well as the distributions of distances and eccentricities in Figure~\ref{fig:star_dist}, taking distances and kinematics from \citep{queiroz2023}. While our stellar sample extends from Galactic center, to the outer disk, to the halo, the majority of our stars ($75\%$) are within 3.5 kpc of the sun. Further, $94\%$ of our stellar sample has an eccentricity less than 0.4, indicative of \textit{in situ} origin \citep[e.g.,][]{sales2009}. In this paper, we assume that the \name fits will be consistent across the Galactic disk, as the median high-Ia and low-Ia [X/Mg] vs. [Mg/H] abundance tends are insensitive to Galactic location (W19, \citealp{griffith2021a}).  While we fit a large sample of stars in this work, \name performs similarly when fitting smaller populations, down to 500 stars.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{star_sample.pdf}
    \caption{Left: distribution of our stellar sample in $Z$ (kpc) vs. $R$ (kpc) where (0,0) is the Galactic center. Center: distribution of stellar distances (kpc). Right: distribution of stellar eccentricity. Our stellar sample spans the Galactic disk, but the majority of our stars are within 3.5 kpc of the sun and have kinematics consistent with \textit{in situ} origin.}
    \label{fig:star_dist}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{all_param1.pdf}
    \caption{Abundance distributions and \name{} parameters for C+N, $\alpha$, and light odd-$Z$ elements. First column: observed abundance distributions in [X/Mg] vs. [Mg/H]. Second column: The predicted [X/Mg] vs. [Mg/H] abundance distribution of the fiducial model plus estimated noise. By comparing the first two columns we can evaluate the success of \name{} in reproducing the observed abundance distributions. Third column: process vector components $\qcc$ (thin, purple) and $\qIa$ (thick, orange) from this work (G23, solid, dark lines) and W22 (light, dashed lines). Overall offsets between the solid and dashed lines are driven largely by our normalization that places the [Mg/Fe] plateau at $+0.4$, rather than $+0.3$ in W22. Fourth column: distribution of fractional contribution from the prompt process ($\fcc$) predicted by the fiducial model. We plot the median $\fcc$ values of the low-Ia (orange square) and high-Ia (purple circle) populations in the solar metallicity bin from W22 for comparison. All density plots are logarithmically scaled.}
    \label{fig:all_param1}
\end{figure*}

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{all_param2.pdf}
    \caption{Same as Figure~\ref{fig:all_param1}, but for Fe-peak elements and Ce.}
    \label{fig:all_param2}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{As.pdf}
    \caption{Left: distribution of $\AIa/\Acc$ vs. $\Acc$ for the fiducial model. This plot is similar to a [Fe/Mg] vs. [Mg/H] distribution where $\AIa/\Acc$ is a proxy for [Fe/Mg] and $\Acc$ is a proxy for [Mg/H]. Note that $\Acc = 1$ corresponds to $\mgh = 0$. Center: distribution of $\AIa/\Acc$ vs. $\mgh$. In the left and center panels, we can clearly see the bimodality to high values of $\Acc$ and $\mgh$. Both density plots are logarithmically scaled. Right: distribution of $\AIa/\Acc$ for ranges of [Mg/H] with metallicity bin increasing from top to bottom. [Mg/H] bins are of width 0.325 dex and span $-0.75$ to 0.55.}
    \label{fig:As}
\end{figure*}

@@

Overall, the fiducial model successfully reproduces the observed abundance distributions. It is capable of capturing metalicity dependences and bimodality. The predicted abundances plus estimated noise are not, however, able to reproduce the observed abundance scatter. This is especially noticeable for O, C+N, Na, Al, K, Co, and Ce. For these elements the scatter in the observed abundance distribution is larger than in the predicted distribution, suggesting that the APOGEE observational scatter is underestimated, that there are $\teff$ or $\logg$ dependent abundance trends (e.g., \citealp{griffith2021a}, W22), or that the $K=2$ model is insufficient--a likely case for elements produced by AGB stars, such as C+N and Ce. The model performs similarly well when the population is downsampled to 5000, 1000, and 500 stars, though the number of nodes has to be decreased from 11 to 7.

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{comp_w22.pdf}
    \caption{Left: cumulative distribution of $\log_{10}(\chi^2)$ for W22 (dashed orange line) and our fiducial model (G23, solid purple line). Right: $\chi^2$ per element for the same model fits with elements ordered by atomic number. Overall, our model has a smaller cumulative $\chi^2$ than the previous two-process model and better predicts the abundances of C+N, Na, K, Mn, Co, Ni, and Ce.}
    \label{fig:comp_w22}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{qccFe_FeMg.pdf}
    \caption{Minimum [Fe/Mg] value attainable for $\qccFe$ and $\dqccFe$ assumptions. Left: for $\dqccFe=0$ with $\qccFe=0.5$, 0.45, 0.4, and 0.35 (light to dark purple). Right: for $\dqccFe=0.15$ with $\qccFe=0.5$, 0.4, 0.35 (light to dark orange).}
    \label{fig:qccFe_FeMg}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{qccFe_FeMgpred.pdf}
    \caption{Top: observed [Fe/Mg] vs. [Mg/H] abundance distribution. Bottom: predicted [Fe/Mg] vs. [Mg/H] abundance distributions plus estimated noise for models with $\dqccFe=0.0$ (middle row) and $\dqccFe=0.15$ (bottom row). The $\qccFe$ value increase from left to right from 0.35 to 0.5. All density plots are logarithmically scaled. Note that the center panels have different $\qccFe$ values (middle: 0.45, bottom: 0.4) and that the predicted abundance distribution for $\qccFe=0.4$ and $\dqccFe=0.0$ can be found in Figure~\ref{fig:all_param2}. The $\qccFe$ and $\dqccFe$ parameters are listed in each panel. By comparing the lower panels to the observed distribution (top) we can evaluate the success or failure of each model; the low-Ia population is most sensitive to the Fe assumptions.}
    \label{fig:qccFe_FeMgpred}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{qccFe_fcc.pdf}
    \caption{Elemental median values of $\fcc$ at solar metallicity for the low-Ia population for \name{} with $\qccFe=0.35$ (darkest purple) to 0.5 (lightest purple) and $\dqccFe=0.0$. Elements are ordered by atomic number. The median $\fcc$ changes most dramatically for elements with strong delayed contributions.}
    \label{fig:qccFe_fcc}
\end{figure*}

@@

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{A4s.pdf}
    \caption{$A^k_i/\Acc$ vs. $\Acc$ distribution for $\AIa$ (left), $A^3_i$ (center), and $A^4_i$ (right). All density plots are logarithmically scaled. The third and fourth processes contribute most to low metallicity stars.}
    \label{fig:A4}
\end{figure*}

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{all_param_k4.pdf}
    \caption{Left: [X/Mg] vs. [Mg/H] abundance distributions for the observed sample (first column), $K=2$ model (second column) and $K=4$ model (third column) for Ca, C+N, Na, Cr, Ni, Mn, Co, and Ce. Observational errors are \textit{not} added to the model predictions. All density plots are logarithmically scaled. Right: process vector components (fourth column) and median low-Ia fractional contribution from each process (fifth column) for the $K=4$ (solid lines) and $K=2$ (dashed lines) models to the low-Ia population as a function of $\mgh$. We plot the median $\qcc$ and $\fcc$ in light purple, $\qIa$ and $f^{\rm Ia}_{ij}$ in dark purple, $q^Z_{3,i}$ and $f^{3}_{ij}$ in light orange, and $q^Z_{4,i}$ and $f^{4}_{ij}$ in dark orange. In the final column we plot a dotted grey line at 0 for reference.} 
    \label{fig:all_paramK}
\end{figure*}
